Debugging by \textit{lastChange}

### 2010-10-13 papers/pldi2011/sigplanconf-template.tex

\begin{abstract}
Backward search from bug symptoms to defects-which cause the bug-is a natural way to debugging. However, due to the lack of support for easily bakward movement by  traditional debugging approaches (e.g., breakpoint-based or log-based debugging), applying this strategy demands a lot of effort from developers. A fundumental step in this process is tracking origins to wrong values and any aid in this regard can greatly reduce the complexity of debugging for developers.

In this paper we present a new functionality in debuggers, \textit{lastChange}, which locates the last place a value is changed and lets developer to collect data and ask for more \textit{lastChange} queries at the point. The only prerequisite for \textit{lastChange} functionality is bug reproduciblity (i.e., there is a test case which reproduces the bug). We explain the algorithms behind this functionality. As a proof of concept we developed a javascript prototype which adds this feature to a well-known javascript debugger, Firebug.

@@

Locating defects cause the bug is the main part of debugging. A common strategy for locating defects is starting from bug symptoms and backward movement by following the origings to wrong values. While this strategy seems straight-forward and effective, applying this strategy using traditional methods is complicated. There are two traditional approaches to debugging, breakpoint-based and log-based debugging. None of these approaches assist developer in finding origings to a wrong value. It is due to developer to search source files, find the list of possible origings to a wrong value and set breakpoints or insert log statements in a way that covers all possible origins.
The next dilemma is data collection. In breakpoint debugging, developer has to memorize values or collect data manually at every breakpoint hit and in cases that there are many breakpoint hits, this becomes a tedious task. In log-based debugging, developer has to decide about data should be collected when inserts the log statement. In this approach developer ends up in dealing with long log files and analyzing huge collected data.

@@

Our contritution in this paper is the algorightm \textit{lastchange}, which locates the last place a value is changed, gathers other values from that execution point, and allows \textit{lastChange} operations from that point. The algorithm builds on existing breakpoint debugger technology. We demonstrate the feasibiltily of the algorithm with an implementation extending the Firebug Javascript debugger.

@@

%---------------------------------------------------------------------------------------------------
\section{Introductory example}
We illustrate the \textit{lastChange} functionality by a simple example. The example demonstrates a buggy javascript code in a html page (Figure ~\ref{fig:js-code}). The page contains a button (line 40) which shows the \texttt{myObject.myProperty} value. Whenever user clicks on the button \texttt{onClick} function is called. This function increases the value of \texttt{myObject.myProperty} by one (line 15) and calls \texttt{updateButton} function which updates button's text to the new value. Once the page is loaded for the first time the button shows \texttt{1} as the initial value of \texttt{myObject.myProperty}. The bug happens when user clicks on the button and instead of \texttt{2}, \texttt{0} appears on the button.

\begin{figure}[htp]
\begin{verbatim}
1 <html>
...
5   <script type="text/javascript">
6    myObject = {myProperty : 1};
7    myCondition = {value : 1};
...
13   function onClick(){
14     foo();
15     myObject.myProperty++;
16     bar();
17     ...
18     updateButton();
19   }
20   function updateButton(){
21     var myParagraph =
          document.getElementById("myButton");
22     myButton.innerHTML = myObject.myProperty;
23   }   
24   function foo(){
25  	 myCondition.value = oldValue;
26   }  
27   function bar(){ 
28     if (!myCondition.value)
29         myObject.myProperty = 0;
30   }
31  </script> 
...
40  <button id="myButton" onclick="onClick()">
41  	1 
42  </button>
43 </html>
\end{verbatim}
\caption{The buggy javascript code.}
\label{fig:js-code}
\end{figure}

To start debugging, the developer sets a abreakpoint on line 22 which is the place button's text is updated. Once the button is clicked execution is paused at line 22. (Figure ~\ref{fig:example1}) shows the debugger while the execution is paused. The developer can examine the program state at the right panel and as it shows the \texttt{myObject.myProperty} value at this point is zero. This value is wrong and the developer seeks the last place this value is changed. By right clicking on \texttt{myObject.myProperty} in the right panel, the developer can run \textit{lastChange} command (Figure ~\ref{fig:example2}). Debugger considers the paused point at the breakpoint as the \textit{reproduction point}, and re-executes the buggy execution, pauses at the reproduction point and shows line 29 in a new panel (Figure ~\ref{fig:example3}). The program state debugger shows on the right panel belongs to the past. Therefore the right panel just shows the partially collected data. If the developer needs some values which are not available in the right panel, debuggers re-executes another time and collects and shows the requested data.

@@

@@

%---------------------------------------------------------------------------------------------------
\section{\textit{lastChange} implementation}
\textit{lastChange} can be called on two different types of values: An object property or a variable value. We explain each case separately.

\subsection{\textit{lastChange} on object propery}
At the beginning, we assume that every object has a unique id. Moreover, we assume that there is a special watchpoint, \textit{property watchpoint}, that can be set on a property name \texttt{foo} and whenever property \texttt{foo} in any object changes this watchpoint hits. We know that none of these assumptions is true in javascript. We discuss this issue later in traking object creation.

Once developer asks for the last change of \texttt{bar.foo} at a reproduction point, debugger sets a property watchpoint on \texttt{foo} and  re-executes the program. Whenever the property watchpoint hits, debugger keeps the object id and stack frame locations. Whenever the execution reaches the same reproduction point it looks at the list of all changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point(Figure ~\ref{fig:foo-changes1}).

@@

\subsection{\textit{lastChange} on variable} 
Once developer asks for the last change of variable \texttt{foo} at a reproduction point, debugger first finds the scope the variable comes from. Consider that in javascript, available variables are not resricted to local varialbes, but all variables in outer scopes are also accessible. The outer scopes are regualr javascript objects or closure scopes. The first group are usually created by calling \texttt{with()} function. The second group are created once a function is called. Therefore, there are theree cases for the escope of a variable, the current scope, an outer scope regular object or an outer scope clousure scope.

@@

%%% we need a short explanation about scopes
%%% important: think about it: we also need to set a special variable (id) to let us to identify the related functional scope
\begin{figure}[htp]
\begin{verbatim}
10  function parent(var1){
11    var var2;
12    var2 = 1;
13    function myfun(var3){
14      var var4;
15      var4 = {...}
16      var2 = var4;
17      function firstChild(var5){
18        var2 = 5;
19        var4 = 5;
20      }  
21      function secondChild(var6){
22       	var var2;
23        var2 = 5			      
24        var4 = 5;	
25      }
26    }  
23  }    
\end{verbatim}
\caption{Sample javascript code demonstrating closure variables.}
\label{fig:js-closure}
\end{figure}

@@

%---------------------------------------------------------------------------------------------------
\section{Javascript implementation}
The javascript prototype is implemented as an extension to Firebug[] javasrcipt debugger.

@@

\subsection{Tracking object creation}
Now we discuss two mentioned assumptions. In the standard javascript every object has a \texttt{watch()} function which receives two parameters, a property name and another function as a hook. Whenever the property with the given name changes the hook function is called. The issue with using this function is that it can only be applied to available objects. However, at the beginning of re-execution there is no object available. To solve this issue we use a feature in Firefox javascript engine, setting a flag gives us the place the object created (e.g., myFile.js, line 24). Though this is a limited data it can help us to get a reference to the object once it is created. Therefore, at the reproduction point, debugger gets the creation location of \texttt{bar} and once the execution reaches to that location it gets a reference to the created object and sets a unique id for it.

>>>>

\begin{abstract}
Backward search from bug symptoms to defects-which cause the bug-is a natural way to debugging. However, due to the lack of support for easily bakward movement by  traditional debugging approaches (e.g., breakpoint-based or log-based debugging), applying this strategy-if it is not pracatically impossible-demands a lot of effort from developers. A fundumental step in this process is tracking origins to wrong values and any aid in this regard can greatly reduce the complexity of debugging for developers.

In this paper we present a new functionality in debuggers, \textit{lastChange}, which locates the last place a value is changed and lets developer to collect data and ask for more \textit{lastChange} queries at the point. The only prerequisite for \textit{lastChange} functionality is bug reproduciblity (i.e., there is a test case which reproduces the bug). We explain the algorithms behind this functionality.

As a proof of concept we developed \textit{RevNav}, a JavaScript prototype, which adds the \textit{lastChange} feature to Firebug JavaScript debugger. Moreover, \textit{RevNav} provides mechanism for automated execution reproduction, and a centralized user interface for all collected results.

@@

Locating defects cause the bug is the main part of debugging. A common strategy for locating defects is starting from bug symptoms and backward movement by following the origings to wrong values. While this strategy seems straight-forward and effective, applying this strategy using traditional methods is complicated. There are two traditional approaches to debugging, breakpoint-based and log-based debugging. None of these approaches assist developer in finding origings to a wrong value. It is due to developer to search source files, find the list of possible origings to a wrong value and set breakpoints or insert log statements in a way that covers all possible origins.

The next dilemma is data collection. In breakpoint debugging, developer has to memorize values or collect data manually at every breakpoint hit and in cases that there are many breakpoint hits, this becomes a tedious task. In log-based debugging, developer has to decide about data should be collected when inserts the log statement. It is very common that the data developer decided to be collected is insufficient. Developer has to change the log statements and re-executes the program another time. Once the sufficient data is collected, it still requires analysis and understanding. Developer usually ends up in dealing with long log files and analyzing huge collected data.

@@

Our contribution in this paper is the algorightm \textit{lastchange}, which locates the last place a value is changed, gathers other values from that execution point, and allows \textit{lastChange} operations from that point. The algorithm builds on existing breakpoint debugger technology. We demonstrate the feasibiltily of the algorithm with an implementation extending the Firebug JavaScript debugger.

@@

%---------------------------------------------------------------------------------------------------
\section{Introductory example}
We illustrate the \textit{lastChange} functionality by a simple example. The example demonstrates a buggy JavaScript code in a html page (Figure ~\ref{fig:js-code}). The page contains a button (line 40) which shows the \texttt{myObject.myProperty} value. It is supposed that whenever user clicks on the button, the value of \texttt{myObject.myProperty} and therefore button's label is increased by one. Once the page is loaded for the first time the button shows \texttt{1} as the initial value of \texttt{myObject.myProperty}. The bug happens when user clicks on the button and instead of \texttt{2}, \texttt{0} appears.

\begin{figure}[htp]
\begin{verbatim}
1 <html>
...
5   <script type="text/javascript">
6    myObject = {myProperty : 1};
7    myCondition = {value : 1};
...
13   function onClick(){
14     foo();
15     myObject.myProperty++;
16     bar();
17     ...
18     updateButton();
19   }
20   function updateButton(){
21     var myParagraph =
          document.getElementById("myButton");
22     myButton.innerHTML = myObject.myProperty;
23   }   
24   function foo(){
25  	 myCondition.value = oldValue;
26   }  
27   function bar(){ 
28     if (!myCondition.value)
29         myObject.myProperty = 0;
30   }
31  </script> 
...
40  <button id="myButton" onclick="onClick()">
41  	1 
42  </button>
43 </html>
\end{verbatim}
\caption{The buggy JavaScript code in a html page.}
\label{fig:js-code}
\end{figure}

Once user clicks on the button, \texttt{onClick} function is called. This function increases the value of \texttt{myObject.myProperty} by one (line 15) and calls \texttt{updateButton} function which updates button's text to the new value (line 22). To start debugging, the developer sets a abreakpoint on line 22 and once the button is clicked execution is paused at line 22. (Figure ~\ref{fig:example1}) shows the debugger while the execution is paused. The developer can examine the program state at the right panel and as it shows the \texttt{myObject.myProperty} value at this point is zero which is obviously a wrong value.

To apply backward search strategy for locating defects, the developer first needs to know the origin to the wrong value. To achieve this goal using breakpoints, the developer should search code to find and set breakpoint at all possible places that \texttt{myObject.myProperty} might get a new value. Even if the developer could sucessfully finish this step (There is no simple way to check it), the next step is monitoring the re-execution and collecting data at breakpoint hits before reaching line 22. All this work is useless if an origin was missed in the first step or if the data was collected in the second step is insufficient.

@@

In contrast, we propose a high-level function in debugger, \textit{lastChange}, which provides the answer without demanding anymore effort from the developer. By right clicking on \texttt{myObject.myProperty} in the right panel, the developer can run \textit{lastChange} command (Figure ~\ref{fig:example2}). Debugger considers the paused point at the breakpoint as the \textit{reproduction point}, and re-executes the buggy execution, pauses at the reproduction point and shows line 29 in a new panel (Figure ~\ref{fig:example3}). The program state debugger shows on the right panel belongs to the past. Therefore the right panel just shows the partially collected data. If the developer needs some values which are not available in the right panel, debugger re-executes another time and collects and shows the requested data.

@@

%---------------------------------------------------------------------------------------------------
\section{\textit{lastChange} Algorithm}
\textit{lastChange} can be called on two different types of values: An object property or a variable value. A variable can be a global, local or closure variable. We explain each case separately.

\subsection{\textit{lastChange} on object propery}

At the beginning, we assume that every object has a unique id. Moreover, we assume that there is a special watchpoint, \textit{property watchpoint}, that can be set on a property name \texttt{foo} and whenever property \texttt{foo} in any object changes this watchpoint hits. We know that none of these assumptions is true in JavaScript. We discuss this issue later in tracking object creation.

Once developer asks for the last change of \texttt{bar.foo} at a reproduction point, debugger sets a property watchpoint on \texttt{foo} and  re-executes the program. Whenever the property watchpoint hits, debugger keeps the object id and stack frame locations. Whenever the execution reaches the reproduction point it looks at the list of all changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point(Figure ~\ref{fig:foo-changes1}).

@@

\subsection{\textit{lastChange} on variable} 
Once developer asks for the last change of variable \texttt{foo} at a reproduction point, debugger first determines the variable's scope as follows:  Consider that in JavaScript, available variables are not resricted to local varialbes, but all variables in outer scopes are also accessible. The outer scopes are regualr JavaScript objects or closure scopes. The first group are usually created by calling \texttt{with()} function. The second group are created once a function is called. Therefore, there are theree cases for the escope of a variable, the current scope, an outer scope regular object or an outer scope clousure scope.

@@

%%% we need a short explanation about scopes
%%% important: think about it: we also need to set a special variable (id) to let us to identify the related functional scope
\begin{figure}[htp]
\begin{verbatim}
10  function parent(var1){
11    var var2;
12    var2 = 1;
13    function myfun(var3){
14      var var4;
15      var4 = {...}
16      var2 = var4;
17      function firstChild(var5){
18        var2 = 5;
19        var4 = 5;
20      }  
21      function secondChild(var6){
22       	var var2;
23        var2 = 5			      
24        var4 = 5;	
25      }
26    }  
23  }    
\end{verbatim}
\caption{Sample JavaScript code demonstrating closure variables.}
\label{fig:js-closure}
\end{figure}

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
The JavaScript prototype is implemented as an extension to Firebug[] javasrcipt debugger.

@@

\subsection{Tracking object creation}
Now we discuss two mentioned assumptions. In the standard JavaScript every object has a \texttt{watch()} function which receives two parameters, a property name and another function as a hook. Whenever the property with the given name changes the hook function is called. The issue with using this function is that it can only be applied to available objects. However, at the beginning of re-execution there is no object available. To solve this issue we use a feature in Firefox JavaScript engine, setting a flag gives us the place the object created (e.g., myFile.js, line 24). Though this is a limited data it can help us to get a reference to the object once it is created. Therefore, at the reproduction point, debugger gets the creation location of \texttt{bar} and once the execution reaches to that location it gets a reference to the created object and sets a unique id for it.
